SignalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-20 paper.tex
finished!

\author[1]{Micha\l{}  Burdukiewicz}
\author[2]{Piotr Sobczyk}
\author[1]{Pawe\l{} B\l{}a\.{z}ej}
\author[1]{Pawe\l{} Mackiewicz}
\affil[1]{University of Wroc\l{}aw, Department of Genomics, Poland}
\affil[2]{Wroc\l{}aw University of Technology, Department of Mathematics, Poland}

@@

\begin{abstract}
signalHsmm using universal, transparent and extendable model of signal peptides, predicts their presence in eukaryotic proteins.
Signal peptides play important role in cell biology as they are responsible for targeting of proteins to endomembrane system and their export outside the cell. Proteins equipped with signal peptides play crucial roles in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. They may be recognized with adequate accuracy using learning systems, that are usually 'black-box' models. Without an ability to trace decision rules, researchers cannot discover which parameters are responsible for predictions. Therefore, we designed a new, more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability. The proposed approach is based on hidden semi-Markov models (HSMMs) and uses intrinsic knowledge about signal peptides. The big advantage of the algorithm is its extensibility. By using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides. Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even though it was trained on very small data sets. Moreover, it proved to be very stable regardless of types of learning data sets. Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases. It should be emphasized, that our model is able to recognize signal peptides from medically significant malaria parasites \textsl{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84). The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.

@@

Despite the low sequence homology between signal peptides~\citep{1999ladungaphysean}, some general architecture were proposed~\citep{1994izardsignal, 2013vossmechanism} - Fig. \ref{fig:sparch}. It is assumed that signal peptides start with a positively charged sequence of amino acid residues, called the n-region with the length of about 5-8 residues. They probably enforce a proper topology on the polypeptide during its translocation through membrane based on the positive-inside rule~\citep{1988vonheijnetopogenic}. The first region is followed by a stretch of hydrophobic amino acids (h-region) with the length of about 8-12 residues. It constitutes a core region of signal peptide and usually forms $\alpha$-helix. The third part of a signal peptide is a polar, but uncharged c-region. It is usually 6 residues long and ends with a cleavage site, in which the signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}. The cleavage site is characterized by a variable amino acid composition. It typically contains small and neutral residues at -3 and -1 positions~\citep{1994palzkillselection}. This site is, however, absent from some membrane proteins in which the first transmembrane domain acts both as a signal peptide and signal anchor~\citep{1988szczesnaskorupapositive}. The amino acid composition and the length of these regions vary between signal peptides, which influences the efficiency of protein secretion~\citep{2006hegdethe}.

\begin{figure}[ht]\centering
\includegraphics[width=0.55\textwidth]{figures/SP.png}
\caption{The organization of typical signal peptide. The figure is not drawn in scale.}
\label{fig:sparch}
\end{figure}

@@

Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious. Therefore, signal peptides became the subject of many computational programs to their prediction. Many software incorporates ‘black-box’ models, such as: neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}. However, these models do not provide direct biological information about organization of signal peptides and are not able to predict properly atypical signal peptides. Although there are programs that do not share the innate flaws of ‘black-box’ models, they also demand an improvement. Some of them are based on position matrices or their variants~\citep{2014zhangprediction, 2004hillerpredisi}. Others (Phobius, Philius and SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which try to reflect structure of signal peptides regions in their limited probabilistic frameworks. The HMMs they use, however, imply a geometric distribution for duration of regions length. We studied the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the the length distribution for every region was not geometric (Fig. \ref{fig:reglen}). Moreover, the commonly used rigid scheme of signal peptide's organization (Fig. \ref{fig:sparch}) does not describe extremely long or short peptides. Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures, but such probabilistic frameworks have not yet been implemented.

@@

All programs used in signal peptide recognition are trained on real protein sequences. Therefore, they succeed in the recognition of peptides similar to those in the learning set but fail in the case of artificial signal peptides. Such peptides are designed to increase effectiveness of protein secretion~\citep{2010futatsumorisugaisignal}. They are especially important in industrial applications to increase yield of proteins. Therefore, only explicit knowledge about the organization of signal peptides allows creating sequences that will be the most efficient in the export of proteins~\citep{2013ngengineering}. Signal peptides have also an important application in gene therapy. Mimicking the natural mechanism of protein export, artificial signal peptides with tumor epitopes increase the antitumor immune response~\citep{2003heenhanced}. Such epitopes must be properly inserted into a signal peptide without decreasing its secretion properties through disruption of the regional structure. Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey in silico many artificial peptides to select the ones that would fulfill the designed role.

Majority of the signal peptide predicting software uses the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid. This method of encoding, however, does not take into account relationships between amino acids and differences in their physicochemical properties. This is disadvantage of such signal peptides' models because their regions are in fact characterized by specific features of amino acid residues and not by the simple occurrence of specific amino acids. In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}. Therefore, we elaborated a new approach based on hidden semi-Markov models using grouping of amino acids into physicochemical groups characteristic of signal peptides. The new methods proved better in comparison to the current software.

@@

Since the functionality of signal peptides depends on the physicochemical properties of residues in a given region, we clustered amino acids into several groups based on their characteristics. The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions~\cite{1998nielsenprediction}. We refined some region recognition criteria to attune the algorithm to less typical signal peptides. Next, two models were trained to recognize proteins with and without a signal peptide. The first one was a hidden semi-Markov model, in which each of three signal peptide regions was represented by a different hidden state. The additional fourth hidden state represented a mature protein. Each state was described by its frequencies of amino acid groups. The distribution of hidden states durations, i.e. the number of amino acids, was based on the empirical distribution of region lengths from the training set. Furthermore, the hidden semi-Markov model was enriched with n-grams representing signal peptide cleavage sites. The second model was a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.

@@

Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06. The positive set contained 2~589 sequences with an experimentally confirmed signal peptide, containing its start and cleavage site information. Sequences with more than one cleavage site were excluded from the final data set. The negative set comprised 152~272 sequences without any signal peptide annotation. Protein sequences with ambiguous symbols: X, J, Z, B and selenocysteine (U) were removed from the final sets.

@@

SignalHsmm is not the first software which uses amino acid encoding in signal peptide prediction. BLOMAP~\citep{maetschke2005blomap} also employed the similar strategy but considered only substitution matrices. However, we applied different approach. We clustered amino acids using four properties relevant for the architecture of signal peptide: their hydrophobicity, frequency in $\alpha$-helices, polarity and size. The high hydrophobicity is determinant of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region. The high polarity as well as small size are important features of residues in the cleavage site~\citep{1994palzkillselection}.

@@

We considered in total 13 amino acid scales from AAIndex database~\citep{2008kawashimaaaindex} (Tab.~\ref{tab:aaprop}). We selected one scale per property and carried out all possible 96 permutations of them. Based on that, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method. Next, we cut the clusterings to create four group of amino acids. In 31\% of cases, the groupings were identical. To compare the usefulness of encodings, we performed a 5-fold cross-validation training a new instance of signalHsmm on every encoding. We created balanced data sets by subsampling proteins without a signal peptide to equal the number of proteins with a signal peptide. The cross-validation was repeated 60 times to ensure that every protein without signal peptide was included in the learning set with probability higher than 0.5. Very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas})  confirms credibility of cross-validation.

\subsection*{Hidden semi-Markov model}
Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM). 
Let us first briefly describe the idea of the HMM. 
Suppose we have a sequence of observations, e.g. amino acids, and we are interested in understanding an underlying cause of their occurrence. 
HMM aims to answer that question assuming a specific and yet flexible structure of the problem.
HMM consists of two stochastic processes. The first is a discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
They are "the cause" of the observations. At every step $t$, hidden state might change according to a transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application, hidden states are signal peptide regions.
The second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$,. They are assumed to occur independently conditionally on the hidden state which emits it. 
Their distribution probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (encoded) amino acids.
The main goal in signalHsmm is to find for a given peptide most probable regions boundaries. This is achieved with Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In the regular HMM, the hidden state duration, i.e. the number of observations emitted by the hidden state, has a geometric distribution. \citep{Durbin98biologicalsequence} showed how to extend it for different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example by~\citep{2004klla}. 
However, it is still not flexible enough because the empirical regional length distributions (see Fig.~\ref{fig:reglen})
are difficult to capture in this way.

%rysunek pokazujący o co chodzi w HMM
\begin{figure}[h]
\centering
\tikzstyle{block} = [draw,shape=circle, top color=white, bottom color=green!50!red!70 ,minimum size=4em]
\tikzstyle{blockComp} = [draw,shape=circle, top color=white, bottom color=red!50!black!70 ,minimum size=2.5em]
% diameter of semicircle used to indicate that two lines are not connected
\def\radius{.7mm} 
\tikzstyle{branch}=[shape=rectangle, top color=white, bottom color=green!50!red!70 ,minimum size=3pt,inner sep=0pt]
\tikzstyle{branch2}=[shape=rectangle, top color=white, bottom color=red!50!black!70 ,minimum size=3pt,inner sep=0pt]
\def\n{11}
 \tikzstyle{line} = [draw=black, color=black!30!white!50, line width=1.5mm, -latex'] 
 \tikzstyle{line2} = [draw=black, color=red!30!black!50, line width=1.5mm, -latex']
\tikzstyle{frame}=[text=black,above, bottom color=white, top color=blue!50!black!70  ]
\tikzstyle{frameComp}=[text=black,below, bottom color=white, top color=red!50!black!70  ]
 \tikzstyle{line3} = [color=green!50!red!70, line width=1.5mm]
\def\names{{"$O_{1,1}$","...","$O_{1,d_1}$", "...","$O_{k, 1}$", "...", "$O_{k, d_k}$"}}%
\begin{tikzpicture}[>=latex']
%wąsate nawiasy
\draw [decorate, decoration={brace,amplitude=10pt},line3,xshift=0pt,yshift=0pt] (2.5,1) -> (10.5,1) node [branch,midway,yshift=0.8cm,color=black] 
	{\textbf{Hidden states}};
% \draw [->,decoration={brace,mirror,amplitude=10pt},line2,xshift=0pt,yshift=0pt] (11,-3) -- (3,-3) node [branch2,midway,yshift=-0.6cm,color=black] 
% 	{\textbf{Kierunek odczytu na nici Cricka ($3' \to 5'$)}}; 
   % Draw blocks, inputs and outputs

\node[block, text width=3em] at (1+2,0) (block1) {\footnotesize 1st hidden state};
\node[blockComp] at (-.5+2,-2)  (komp1) {\footnotesize \pgfmathparse{\names[0]}\pgfmathresult};
\node[blockComp] at (1+2,-2)  (komp2) {\footnotesize \pgfmathparse{\names[1]}\pgfmathresult};
\node[blockComp] at (2.5+2,-2)  (komp3) {\footnotesize \pgfmathparse{\names[2]}\pgfmathresult};
\draw[line,->]  (block1) --  (komp1);
\draw[line,->]  (block1) --  (komp2);
\draw[line,->]  (block1) --  (komp3);
\draw[line3,->] (block1.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};

\node[block, bottom color=green!50!red!80,] at (4.5+2,0) (block2) {...};
\draw[line3,->] (block2.east) -- +(2,0) node [branch,midway,yshift=0.2cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
\node[blockComp] at (4.5+2,-2)  (komp4) {\footnotesize \pgfmathparse{\names[3]}\pgfmathresult};
\draw[line,->]  (block2) --  (komp4) ;

\node[block,text width=3em, bottom color=green!50!red!90] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
\node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};
\node[blockComp] at (8+2,-2)  (komp6) {\footnotesize \pgfmathparse{\names[5]}\pgfmathresult};
\node[blockComp] at (9.5+2,-2)  (komp7) {\footnotesize \pgfmathparse{\names[6]}\pgfmathresult};
\draw[line,->]  (block3) --  (komp5);
\draw[line,->]  (block3) --  (komp6);
\draw[line,->]  (block3) --  (komp7);

@@

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
(1,-1.8) -- (5,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_1$}};

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
(8,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_k$}};

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=15pt,mirror,raise=10pt},yshift=-10pt]
(1,-2.6) -- (12,-2.6) node [branch2, midway, yshift=-1.3cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
\end{tikzpicture}
\caption{General scheme of hidden semi-Markov model.}
\label{fig:hsmm}
\end{figure}

The model we used is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
In addition to matrices $A$ and $B$, the model is given by probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
where $D$ is the maximum allowed duration.
As our datasets are of reasonable size and $D$ is small -- around 30, computational effort is not much 
higher than in the regular HMM.

Our model has a very specific structure. The hidden states represent signal peptide regions.
Almost all entries in the transition matrix $A$ are zeros because regions are sequential.
Possible transitions are depicted as arrows in Fig.~\ref{fig:signalHsmm}.
Probabilities of observations for the hidden states and hidden states durations were estimated from training data.
The advantage of HSMM model is not only a better performance but also its straightforwardness.
Fig.~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.

%model structure
\begin{figure}
 \begin{minipage}{.5\linewidth} 
     \resizebox{6.5cm}{!}{%
    \begin{tikzpicture}[->,>=stealth',shorten >=2pt,auto,node distance=4.5cm, thick]
    \label{fig:model_hsmm}
      \tikzstyle{line} = [draw=black, color=blue!30!black!50, line width=4.5mm, -latex']
      \tikzstyle{main node} = [circle,fill=blue!20,draw, minimum size = 2.2cm, font=\itshape,
         align=center,  top color=white, bottom color=blue!50!black!70 ] %font=\sffamily\small\bfseries,
      %nodes
      \node[main node]            (start') 	[]						{Start};	     
      \node[main node, bottom color=purple!70!black!70] 	(nregion') 	[right of=start',xshift=-5mm, yshift=15mm] 	{n-region};
      \node[main node, bottom color=pink!70!black!70] 	(hregion') 	[right of=nregion',xshift=-5mm,yshift=15mm] 	{h-region};
      \node[main node, bottom color=gray!70!black!70] 	(cregion') 	[right of=hregion',xshift=-5mm,yshift=-15mm] 	{c-region};
      \node[main node, bottom color=green!70!black!70] 	(mature') 	[right of=cregion',xshift=-5mm, yshift=-15mm] 	{Mature protein};

>>>>

\author[1]{Micha\l{}  Burdukiewicz}
\author[2]{Piotr Sobczyk}
\author[1]{Pawe\l{} B\l{}a\.{z}ej}
\author[1]{Pawe\l{} Mackiewicz}
\affil[1]{University of Wroc\l{}aw, Department of Genomics, Poland; michalburdukiewicz@gmail.com, blazej.pawel@gmail.com, pamac@smorfland.uni.wroc.pl}
\affil[2]{Wroc\l{}aw University of Technology, Department of Mathematics, Poland; piotr.sobczyk@pwr.edu.pl}

@@

\begin{abstract}
Signal peptides play an essential role in targeting of proteins to endomembrane system and their export outside the cell. The proteins equipped with such peptides are of great importance in metabolism, maintenance of tissue structure, immune response and regulation of other organismal functions. These peptides may be recognized with adequate accuracy using learning systems. However, they are usually 'black-box' models therefore unable to trace decision rules and identify parameters responsible for the predictions. That is why, we designed a new more universal probabilistic model for eukaryotic signal peptides, which includes knowledge about their organization, amino acid composition and variability. The proposed approach called signalHsmm is based on hidden semi-Markov models (HSMMs) and uses intrinsic knowledge about signal peptides. The big advantage of the algorithm is its extensibility. By using the n-grams (k-mers), we showed that the general model can yield not only better results than other software but also more information about features of signal peptides. Our model showed the largest AUC = 0.97 in comparison to other signal peptide predictor even though it was trained on very small data sets. Moreover, it proved to be very stable regardless of types of learning data sets. Therefore, our model does not need to be permanently retrained with the continuous expansion of sequence databases. It should be emphasized, that our model is able to recognize signal peptides from medically significant malaria parasites \textit{Plasmodium} and their relatives more accurately (with AUC = 0.92) than popular programs (0.84). The web-server of signalHsmm is available at \url{http://smorfland.uni.wroc.pl/signalhsmm}.

@@

Despite the low sequence homology between signal peptides~\citep{1999ladungaphysean}, some general architecture were proposed~\citep{1994izardsignal, 2013vossmechanism} - Fig. \ref{fig:sparch}. It is assumed that signal peptides start with a positively charged sequence of amino acid residues, called the n-region with the length of about 5-8 residues. They probably enforce a proper topology on the polypeptide during its translocation through membrane based on the positive-inside rule~\citep{1988vonheijnetopogenic}. The first region is followed by a stretch of hydrophobic amino acids (h-region) with the length of about 8-12 residues. It constitutes a core region of signal peptide and usually forms $\alpha$-helix. The third part of a signal peptide is a polar, but uncharged c-region. It is usually 6 residues long and ends with a cleavage site, in which a signal peptidase cleaves the signal peptide, during or after translocation of the protein into the endoplasmic reticulum ~\citep{2002paetzelsignal}. The cleavage site is characterized by a variable amino acid composition. It typically contains small and neutral residues at -3 and -1 positions~\citep{1994palzkillselection}. This site is, however, absent from some membrane proteins in which the first transmembrane domain acts both as a signal peptide and signal anchor~\citep{1988szczesnaskorupapositive}. The amino acid composition and the length of these regions vary between signal peptides, which influences the efficiency of protein secretion~\citep{2006hegdethe}.

\begin{figure}[ht]\centering
\includegraphics[width=0.55\textwidth]{figures/SP.eps}
\caption{The organization of typical signal peptide. The regions are not drawn in scale.}
\label{fig:sparch}
\end{figure}

@@

Although many experimental methods determining the subcellular localization of proteins were devised, they are time consuming and laborious. Therefore, signal peptides became the subject of many computational programs to their prediction. Many software incorporates ‘black-box’ models, such as: neural networks~\citep{2011petersensignalp}, support vector machines~\citep{2014zhangprediction}, Bayesian networks~\citep{2012zhengsignalbnf} or k-nearest neighbours~\citep{2007shensignall}. However, these models do not provide direct biological information about organization of signal peptides and are not able to predict properly atypical signal peptides. Although there are programs that do not share the innate flaws of ‘black-box’ models, they also demand an improvement. Some of them are based on position matrices or their variants~\citep{2014zhangprediction, 2004hillerpredisi}. Others (Phobius, Philius and SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which try to reflect structure of signal peptides regions in their limited probabilistic frameworks. The used HMMs, however, imply a geometric distribution for duration of regions length. We studied the distribution for regions from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the length distribution for every region was not geometric (Fig. \ref{fig:reglen}). Moreover, the commonly used rigid scheme of signal peptide's organization (Fig. \ref{fig:sparch}) does not describe extremely long or short peptides. Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also the unusual structures but such probabilistic frameworks have not yet been implemented.

@@

All programs used in signal peptide recognition are trained on real protein sequences. Therefore, they succeed in the recognition of peptides similar to those in the learning set but fail in the case of artificial signal peptides. Such peptides are designed to increase effectiveness of protein secretion~\citep{2010futatsumorisugaisignal}. They are especially important in industrial applications to increase yield of proteins. Therefore, only explicit knowledge about the organization of signal peptides allows creating sequences that will be the most efficient in the export of proteins~\citep{2013ngengineering}. Signal peptides have also an important application in gene therapy. Mimicking the natural mechanism of protein export, artificial signal peptides with tumor epitopes increase the antitumor immune response~\citep{2003heenhanced}. Such epitopes must be properly inserted into a signal peptide without decreasing its secretion properties through disruption of the regional structure. Instead of time-consuming and expensive laboratory experiments, it would be very useful to survey \textit{in silico} many artificial peptides to select the ones that would fulfill the designed role.

Majority of the signal peptide predicting software uses the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid. This method of encoding, however, does not take into account relationships between amino acids and differences in their physicochemical properties. This is disadvantage of such signal peptide models because their regions are in fact characterized by specific features of amino acid residues and not by the simple occurrence of specific amino acids. In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}. Therefore, we elaborated a new approach based on hidden semi-Markov models using grouping of amino acids into physicochemical groups characteristic of signal peptides. The new methods proved better in comparison to the current software.

@@

Since the functionality of signal peptides depends on the physicochemical properties of residues in a given region, we clustered amino acids into several groups based on their characteristics. The pre-processed sequences were further analyzed by the heuristic algorithm, which determines borders between three characteristic signal peptide regions~\citep{1998nielsenprediction}. We refined some region recognition criteria to attune the algorithm to less typical signal peptides. Next, two models were trained to recognize proteins with and without a signal peptide. The first one was a hidden semi-Markov model, in which each of three signal peptide regions was represented by a different hidden state. The additional fourth hidden state represented a mature protein. Each state was described by its frequencies of amino acid groups. The distribution of hidden states durations, i.e. the number of amino acids, was based on the empirical distribution of region lengths from the training set. Furthermore, the hidden semi-Markov model was enriched with n-grams representing signal peptide cleavage sites. The second model was a simple probabilistic approach in which no association between amino acids was assumed, and probability of amino acids groups occurrence was determined by their frequencies in mature proteins.

@@

Eukaryotic protein sequences and their annotations were properly prepared according to the literature of the subject and downloaded from UniProt database release 2015\_06. The positive set contained 2~589 sequences with an experimentally confirmed signal peptide, containing its start and cleavage site information. Sequences with more than one cleavage site were excluded from the final data set. The negative set comprised 152~272 sequences without any signal peptide annotation. Protein sequences with ambiguous symbols: X, J, Z, B and selenocysteine (U) were finally removed.

@@

SignalHsmm is not the first software which uses amino acid encoding in signal peptide prediction. BLOMAP~\citep{maetschke2005blomap} also employed the similar strategy but considered only substitution matrices. However, we applied different approach. We clustered amino acids using four properties relevant for the architecture of signal peptide: their hydrophobicity, frequency in $\alpha$-helices, polarity and size. The high hydrophobicity is a determinant of the h-region, whose $\alpha$-helix secondary structure is probably induced by the positively charged n-region. The high polarity as well as small size are important features of residues in the cleavage site~\citep{1994palzkillselection}.

@@

We considered in total 13 amino acid scales from AAIndex database~\citep{2008kawashimaaaindex} (Tab.~\ref{tab:aaprop}). We selected one scale per property and carried out all possible 96 permutations of them. Based on that, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method. Next, we cut the clusterings to create four group of amino acids. In 31\% of cases, the groupings were identical. To compare the usefulness of encodings, we performed a 5-fold cross-validation training the new instance of signalHsmm on every encoding. We created balanced data sets by subsampling proteins without a signal peptide to equal the number of proteins with a signal peptide. The cross-validation was repeated 60 times to ensure that every protein without signal peptide was included in the learning set with the probability higher than 0.5. Very small variance of performance measures (for example see Tab.~\ref{tab:perfmeas})  confirmed the credibility of cross-validation.

\subsection*{Hidden semi-Markov model}
Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM). 
Let us first briefly describe the idea of the HMM. 
Suppose we have a sequence of observations, e.g. amino acids, and we are interested in understanding an underlying cause of their occurrence. 
HMM aims to answer that question assuming a specific and yet flexible structure of the problem.
HMM consists of two stochastic processes. The first is a discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
They are "the cause" of the observations. At every step $t$, hidden state might change according to a transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application, hidden states are signal peptide regions.
The second process $E_{t=1}^T$ is an observation process defined on the set of possible observations $\{O_1, \dots, O_m\}$. They are assumed to occur independently and conditionally on the hidden state that emits it. 
Their distribution are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, the observations are (encoded) amino acids.
The main goal in signalHsmm is to find the most probable regions boundaries for a given peptide. This is achieved with Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

In the regular HMM, the hidden state duration, i.e. the number of observations emitted by the hidden state, has a geometric distribution. \cite{Durbin98biologicalsequence} showed how to extend it for different distributions without significant increase in computational 
complexity. Similar ideas were used for signal peptide recognition, for example by~\cite{2004klla}. 
However, it is still not flexible enough because the empirical regional length distributions (see Fig.~\ref{fig:reglen})
are difficult to capture in this way.

%rysunek pokazujący o co chodzi w HMM
\begin{figure}[h]
\centering
\tikzstyle{block} = [draw,shape=circle, top color=green!50!white!70, bottom color=green!50!white!70 ,minimum size=4em]
\tikzstyle{blockComp} = [draw,shape=circle, top color=red!50!white!70, bottom color=red!50!white!70 ,minimum size=2.5em]
% diameter of semicircle used to indicate that two lines are not connected
\def\radius{.7mm} 
\tikzstyle{branch}=[shape=rectangle, top color=white, bottom color=white ,minimum size=3pt,inner sep=0pt]
\tikzstyle{branch2}=[shape=rectangle, top color=white, bottom color=white ,minimum size=3pt,inner sep=0pt]
\def\n{11}
 \tikzstyle{line} = [draw=black, color=black!70!white!50, line width=1.5mm, -latex'] 
 \tikzstyle{line2} = [draw=black, color=red!30!white!70, line width=1.5mm, -latex']
\tikzstyle{frame}=[text=black,above, bottom color=white, top color=blue!50!black!70  ]
\tikzstyle{frameComp}=[text=black,below, bottom color=red!50!white!70, top color=red!50!white!70  ]
 \tikzstyle{line3} = [color=green!50!black!70, line width=1.5mm]
\def\names{{"$O_{1,1}$","...","$O_{1,d_1}$", "...","$O_{k, 1}$", "...", "$O_{k, d_k}$"}}%
\begin{tikzpicture}[>=latex']
%wąsate nawiasy
\draw [decorate, decoration={brace,amplitude=10pt},line3,xshift=0pt,yshift=0pt] (2.5,1) -> (10.5,1) node [branch,midway,yshift=0.8cm,color=black] 
	{\textbf{Hidden states}};
% \draw [->,decoration={brace,mirror,amplitude=10pt},line2,xshift=0pt,yshift=0pt] (11,-3) -- (3,-3) node [branch2,midway,yshift=-0.6cm,color=black] 
% 	{\textbf{Kierunek odczytu na nici Cricka ($3' \to 5'$)}}; 
   % Draw blocks, inputs and outputs

\node[block, text width=3em] at (1+2,0) (block1) {\footnotesize 1st hidden state};
\node[blockComp] at (-.5+2,-2)  (komp1) {\footnotesize \pgfmathparse{\names[0]}\pgfmathresult};
\node[blockComp] at (1+2,-2)  (komp2) {\footnotesize \pgfmathparse{\names[1]}\pgfmathresult};
\node[blockComp] at (2.5+2,-2)  (komp3) {\footnotesize \pgfmathparse{\names[2]}\pgfmathresult};
\draw[line,->]  (block1) --  (komp1);
\draw[line,->]  (block1) --  (komp2);
\draw[line,->]  (block1) --  (komp3);
\draw[line3,->] (block1.east) -- +(2,0) node [branch,midway,yshift=0.3cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};

\node[block, bottom color=green!50!white!80,] at (4.5+2,0) (block2) {...};
\draw[line3,->] (block2.east) -- +(2,0) node [branch,midway,yshift=0.3cm,xshift=-0.2cm,color=black] {\footnotesize \textbf{transition}};
\node[blockComp] at (4.5+2,-2)  (komp4) {\footnotesize \pgfmathparse{\names[3]}\pgfmathresult};
\draw[line,->]  (block2) --  (komp4) ;

\node[block,text width=3em, bottom color=green!50!white!70] at (8+2,0) (block3) {\footnotesize $k^{th}$ hidden state};
\node[blockComp] at (6.5+2,-2)  (komp5) {\footnotesize \pgfmathparse{\names[4]}\pgfmathresult};
\node[blockComp] at (8+2,-2)  (komp6) {\footnotesize \pgfmathparse{\names[5]}\pgfmathresult};
\node[blockComp] at (9.5+2,-2)  (komp7) {\footnotesize \pgfmathparse{\names[6]}\pgfmathresult};
\draw[line,->]  (block3) --  (komp5);
\draw[line,->]  (block3) --  (komp6);
\draw[line,->]  (block3) --  (komp7);

@@

\draw [color=red!50!black!70, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
(1,-1.8) -- (5,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_1$}};

\draw [color=red!50!black!70, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
(8,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_k$}};

\draw [color=red!50!black!70, line width=1.5mm, decorate,decoration={brace,amplitude=15pt,mirror,raise=10pt},yshift=-10pt]
(1,-2.6) -- (12,-2.6) node [branch2, midway, yshift=-1.3cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};
\end{tikzpicture}
\caption{General scheme of hidden semi-Markov model.}
\label{fig:hsmm}
\end{figure}

The model we used is Hidden semi-Markov Model (HSMM)~\citep{Yu2010215}. It extends HMM by allowing any given hidden state duration distribution (Fig.~\ref{fig:hsmm}).
In addition to matrices $A$ and $B$, the model is given by probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
where $D$ is the maximum allowed duration.
As our datasets are of reasonable size and $D$ is small -- around 30, computational effort is not much 
higher than in the regular HMM.

Our model has a very specific structure. The hidden states represent signal peptide regions.
Almost all entries in the transition matrix $A$ are zeros because regions are sequential.
Possible transitions are depicted as arrows in Fig.~\ref{fig:ngramext}.
Probabilities of observations for the hidden states and hidden states durations were estimated from training data.
The advantage of HSMM model is not only a better performance but also its straightforwardness.
Fig.~\ref{fig:ngramext} is easy to interpret for a researcher without any mathematical background.

%model structure

\begin{figure}[ht]\centering
\includegraphics[width=0.51\textwidth]{figures/HSMMs.eps}
\caption{The diagram of simple (A) and extended version of signalHsmm with the n-gram cleavage site model (B).}
\label{fig:ngramext}
\end{figure}
